The hidden manifold distance for functional data

### 2019-07-29 index.Rmd
more writing of intro and sim

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that, since observations are smooth over their domain, the dimensionality of the curves is only artifically high.
Less studied in functional data analysis is the idea that the curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include the class of probability density functions and the class of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of calculating the pairwise geodesic distance between functional observations that, due to noise, live near a non-linear manifold rather than exactly on the manifold. Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis.
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
We conclude the work by using our pairwise geodesic distance estimator in distanced-based functional classification.

# Introduction [Susan]

Let $X_1,\ldots,X_n$ be a sample of $n$ independant realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. We additionally assume that the function $X$ belongs to a low-dimensional nonlinear manifold $\M \subset L^2([a,b],\R)$. In this case a geodesic distance taking into account the intrinsic structure of $\M$ should be used as a measure of distance between two functions in $\M$, rather than the $L^2$ distance. As many statistical methods depend on some measure of distance, downstream benefits of using the geodesic distance can be expected.

@@

Given perfect functional observations, i.e.\ with no measurement error and on a very dense domain grid, the estimation of the geodesic distance can be accomplished straightforwardly using algorithms such as Floyd's. 
However, when functional data are observed with noise, this approach is likely to fail. This is because we assume the error is such that the noisy functional manifold data no longer live on a manifold. 
Though powerful techniques exist for smoothing discretely observed noisy functional data, the smoothing is designed for the $L_2$ recovery of the original curve, not to ensure the recovered functional versions of the data lie on or close to the functional manifold $\M$. 
<!-- This presents challenges to the application of many manifold learning techniques which assume fully observed nearly-noiseless inputs. -->

@@

With this challenge in mind, we put forth a technique for the specific task of estimating pairwise geodesic distances $\{ d_\M (X_i,X_j\}_{i>j}$, to be defined  when we have access only to discretely-observed noisy functional observations that possibly live off the true manifold. Briefly outline the methodology: MDS plus density ridge are used in a cool way to estimate (consistently?, let's hope..) pairwise geodesic distances.

@@

## Related Work [Susan]
[@ChenMuller2012]

@@

## Setting [Susan]
Let $X_1,\ldots,X_n$ be independent realizations of a random function $X\in\M \subset L^2([a,b],\R)$, where $\M$ is an unknown manifold. Suppose that each curve $X_i$ is observed with measurements errors on a grid $T_i=(t_{i1},\ldots,t_{iK})$, i.e. we observe a sample of $K$-dimensional vectors $Y_1,\ldots,Y_n$ with $Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}$, where the random variables $\epsilon_{ij}$ are of mean zero and uncorrelated with each other. We assume that the grids $T_1,\ldots,T_n$ are dense.

@@

## Relevant differential geometry concepts [Susan]

We first rigorously define the concepts of geodesic and geodesic distance. Following [@Lin2014], let $(\M,g)$ be a $d$-dimensional Riemannian manifold, where $g$ is a Riemannian metric tensor on $\M$ which can be used to assign a metric on the manifold. Specifically, for each point $p$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_p$ on the tangent space $T_p \M$. The norm of a tangent vector $v \in T_p \M$ is defined as $$||v|| = \sqrt{g_p(v,v)}.$$
If $\gamma: [a,b] \subset \R \to \M$ is a smooth curve, its length is then defined as
$$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt .$$
The geodesic distance between two points $p,q$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.$$
<!-- ##############  -->

@@

\begin{enumerate}
\item The first step consists in constructing a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\R^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_\M(x_i,x_j)$
are estimated based on $G$. The geodesic distance between two nodes is simply the length of the shortest path in $G$ between these two nodes, which we calculate with the Floyd-Warshall algorithm [@Warshall1962]
<!-- or Dijkstra’s  which one do we actually use in our implementation, the FW or D? -->
\item The last step consists in using MDS on the matrix of pairwise geodesic distances obtained in the first step in order to obtain an embedding in $\R^d$.
\end{enumerate}

@@

We perform a simulation study to ascertain the efficacy of our method for estimating pairwise geodesic distances for discretely-observed noisy functional data.

@@

## Performance metrics [Susan]
We describe three different metrics we use to assess the quality of a pairwise geodesic distance estimator. This is the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.

## Simulation scenarios [Susan]

@@

### Sampling from a functional manifold
As of yet, there is little work as to how to sample from a functional manifold. Even in the Euclidean case, it is not obvious how sampling should be done [@Diaconis2013]. To safeguard against sampling unevenly on the functional manifold, we sample on a very concentrated measure for the intrinsic parameters.
For both the manifold of normal densities and the manifold of warping functions, we sample $\alpha$ according to ???rtruncnorm(samplesize,alphamin,alphamax,(alphamin+alphamax)/2,1)???
For the manifold of square root beta densities, we sample as follows       
alpha<- rtruncnorm(nb_alpha,alphamin,alphamax,(alphamax+alphamin)/2,0.3)
beta<- rtruncnorm(nb_beta,betamin,betamax,(betamax+betamin)/2,0.3)

How to sample properly from a functional manifold could be interesting future work.
### Results

Submit compare_methods.sh as slurm job using sbatch.
```{r, engine = 'bash', eval = FALSE}
sbatch compare_methods.sh
```

@@

Simulation results are then visualized by invoking cluster_results.R. There's some unfortunate hardcoding still in cluster_results.R
```{r, eval=FALSE}
R cluster_results.R
```

@@

>>>>

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that the dimensionality of curves is only artifically high if they are smooth.
There is also another sense in which the dimensionality of the curves under study may be low. This is the much less explored idea that curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include classes of probability density functions and classes of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of estimation the pairwise geodesic distance between functional manifold data when we only have access to their noisy realizations which live near a non-linear manifold rather than exactly on the manifold. 
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis which we illustrate in the case of distanced-based functional classification.

# Introduction 
<!-- Susan  -->
Many statistical methods depend on a measure of distance. Methods in functional data analysis, the study of infinite-dimensional smooth curves, is no exception. In situations where the manifold hypothesis might be plausible, one might consider using a geodesic distance that takes into account the intrinsic structure of the manifold rather than the standard $L^2$ distance. The benefits of adopting the geodesic distance may be realized in downstream tasks such as distanced-based clustering and classification of functional obvervations.

Specifically, let $X_1,\ldots,X_n$ be a sample of $n$ independant realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. Suppose additionally that the function $X$ belongs to a low-dimensional nonlinear Riemannian manifold $\M \subset L^2([a,b],\R)$ where $g$ is a Riemannian metric tensor on $\M$ which can be used to assign a metric on the manifold as follows [@Lin2014]. For each point $X$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_X$ on the tangent space $T_X \M$. The norm of a tangent vector $V \in T_X \M$ is defined as $$||V|| = \sqrt{g_X(V,V)}.$$ The geodesic distance between two functions $X_i,X_j$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(X_i,X_j):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = X_i, \gamma(b) = X_j]  \}$$
where $$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt$$
is the length of a smooth curve $\gamma: [a,b] \subset \R \to \M$.

@@

Given perfect functional observations $X_i$ and $X_j$, i.e.\ with no measurement error and on a very dense domain grid, the estimation of the geodesic distance $d_g(X_i,X_j)$ can be accomplished straightforwardly using a shortest-path algorithm, e.g. the Floyd-Warhsall algorithm [@Floyd1962]. 
However, when functional data are observed with noise, this approach is likely to fail. This is because the error may be such that the noisy functional manifold data no longer live on a manifold. 
A preprocessing step where the nosiy functional manifold data is smoothed is also likely to fail. This is because smoothing techniques in functional data analysis are designed for the $L_2$ recovery of the original curve, not to ensure the recovered functional versions of the data lie on or close to the functional manifold $\M$. 
<!-- This presents challenges to the application of many manifold learning techniques which assume fully observed nearly-noiseless inputs. -->

@@

In this work, we put forth a technique for the specific task of recovering pairwise geodesic distances $\{ d_\M (X_i,X_j)\}_{i>j}$ when we have, in lieu of $X_i$ and $X_j$, access only to discretely-observed noisy functional observations $Y_i$ and $Y_j$ that possibly live off the true manifold. This is accomplished by first projecting the functional data on to the underlying hidden manifold and then performing operations on this hidden manifold.

@@

## Related Work
<!-- Susan  -->
[@ChenMuller2012]

@@

Suppose that each curve $X_i$ is observed with measurements errors on a grid $T_i=(t_{i1},\ldots,t_{iK})$, i.e. we observe a sample of $K$-dimensional vectors $Y_1,\ldots,Y_n$ with $Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}$, where the random variables $\epsilon_{ij}$ are of mean zero and uncorrelated with each other. We assume that the grids $T_1,\ldots,T_n$ are dense.

@@

@@

\begin{enumerate}
\item The first step consists in constructing a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\R^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_\M(x_i,x_j)$
are estimated based on $G$. The geodesic distance between two nodes is simply the length of the shortest path in $G$ between these two nodes, which we calculate with the Floyd-Warshall algorithm [@Floyd1962].
<!-- Which one do we actually use in our implementation, Floyd-Warshall or Dijkstra's? -->
\item The last step consists in using MDS on the matrix of pairwise geodesic distances obtained in the first step in order to obtain an embedding in $\R^d$.
\end{enumerate}

@@

We perform a simulation study to ascertain the efficacy of our method for estimating pairwise geodesic distances when one only has access to discretely-observed noisy functional data. Three different metrics are used to assess the quality of a pairwise geodesic distance estimator.

* Near-$\epsilon$ isometry: area under the curve of the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.  
* Relative Frobenius: let $d$ and $\hat d$ be the true and estimated $n \times n$ geodesic distance matrices, respectively. 
$$
||d-\hat d||/||d||
$$
* Pearson correlation: take the upper diagonal of $d$ and $\hat d$, calculate their Pearson correlation coefficient.

@@

## Simulation scenarios 
<!-- Susan  -->

@@

## Sampling from a functional manifold
As of yet, there is little work as to how to sample from a functional manifold. Even in the Euclidean case, it is not obvious how sampling should be done [@Diaconis2013]. To safeguard against sampling unevenly on the functional manifold, we sample on a very concentrated measure for the intrinsic parameters.
For both the manifold of normal densities and the manifold of warping functions, we sample $\alpha$ according to ???rtruncnorm(samplesize,alphamin,alphamax,(alphamin+alphamax)/2,1)???
For the manifold of square root beta densities, we sample as follows       
alpha<- rtruncnorm(nb_alpha,alphamin,alphamax,(alphamax+alphamin)/2,0.3)
beta<- rtruncnorm(nb_beta,betamin,betamax,(betamax+betamin)/2,0.3)

How to sample properly from a functional manifold could be interesting future work.

## Results
Submit compare_methods.sh as slurm job using sbatch.
```{r, engine = 'bash', eval = FALSE}
sbatch compare_methods.sh
```

@@

Simulation results are then visualized by invoking cluster_results.R. There's some unfortunate hardcoding still in cluster_results.R
```{r, eval=FALSE}
R cluster_results.R
```
<!--

@@

-->
